Unsupervised embeddings of organizations capture latent structure of scientific mobility

### 2020-04-13 paper/current/main.tex

Why does the word embedding model, originally designed for natural language processing, produce a reasonable distance that models gravity relationships between organizations?
We find evidence of a theoretical connection between the two models, one that makes the skip-gram model applicable to embedding many forms of mobility trajectories.

According to our observation on the embedding space,

\begin{equation}
    \label{eq:gravity_model}
    \frac{T_{ij}}{M_{i} M_{j}} \propto -\exp(d_{ij}) = - \exp(1 - \cos(v_i,v_j)) \propto \exp(\cos(v_i,v_j)),
\end{equation}

where $T_{ij}$ represents the \textbf{flux}, the number of people who moved between organizations $i$ and $j$,  $M_{i}$ and $M_{j}$ represent the size of an organization,  $d_{ij} = 1 - \cos(v_{i}, v_{j})$ is the embedding distance, and $\cos(v_{i}, v_{j})=\frac{v_{i} \cdot v_{j}}{\lVert v_{i} \rVert \lVert v_{j} \rVert}$  is the cosine similarity between two vectors. In the case of scientific co-affiliation, we assume all movements are without direction (symmetric), meaning that  $T_{ij} = T_{ji}$.

To convert Eq. \ref{eq:gravity_model} to a flux-related term only, we replace  $M_*$ with $\mu_{i*}$, the total flux to or from $i$ which can be written as

@@

\begin{equation}
    \label{eq:skipgram_log_prob_condensed}
    \frac{1}{T}\sum_{t=1}^{T} \sum_{-c \le j \le c, j \neq 0} \log p(w_{t+j}|w_t),
\end{equation}

where the conditional probability of observing one word, $w_{O}$, given another, $w_{I}$, is defined as,

\begin{equation}
    \label{eq:skipgram_conditional_expanded}
    p(w_O|w_I) = \frac{\exp(v'_{w_O} \cdot v_{w_I})}{\sum_{w=1}^{W} \exp(v'_w \cdot v_{w_I})},
\end{equation}

where $v_w$ and $v'_w$ are ``input'' and ``output'' vector representations of each word in the vocabulary.  We use Skip-gram negative sampling, and following the work by Levy and Goldberg \autocite{levy2014neural}, consider the model as implicitly performing matrix factorization as shown below,

\begin{equation}
    \label{eq:sgns_relation}
    \log \left(\frac{\#(w,c) |D|}{\#(w) \#(c)} \right) - \log k,
\end{equation}

where $\#(w)\text{, } \#(c) \text{ and } \#(w, c)$ denotes the number of times word $w$, $c$, and word-context-pair ($w$, $c$) in the corpus respectively. Variable $k$ is the number of negative samples and $|D|$ is the size of the corpus.

Mathematically, the maximization of the objective function of the model yields $V=L$ and $V'=R$ where L and R are the left and right singular vectors of the matrix Eq.\ref{eq:sgns_relation}, respectively. Note that $V=L=R=V'$ because Eq.\ref{eq:sgns_relation} is a real symmetric matrix. With these relations, for a specific pair of words $(w_i ,w_j )$,

\begin{equation}
    \label{eq:sgns_relation_w,i, w_j}
    v_{w_i} \cdot v_{w_j} =  \log \left(\frac{\#(w_i, w_j) |D|}{\#(w_i) \#(w_j)} \right) - \log k,
\end{equation}

where $|D|$ and $k$ are constants, such that the functional form can be re-written as,

\begin{equation}
    \label{eq:sgns_relation_w,i, w_j}
     \frac{\#(w_i, w_j)}{\#(w_i) \#(w_j)}  \propto  \exp(v_{w_i} \cdot v_{w_j} ).
\end{equation}

In our model, $\#(w_i)$ and $\#(w_j)$ are total flux $M_i$ and $M_j$ respectively,  and $\#(w_i,w_j)$ is just  $T_{ij}$, flux between $i$ and $j$. Then, we have the same expression with gravity model,

>>>>

Why does the word embedding model, originally designed for natural language processing, produce a reasonable distance that models gravity relationships between organizations?
We find evidence of a theoretical connection between the two models, one that makes the neural embedding model applicable to embedding many forms of mobility trajectories.

According to our observation on the embedding space from Eq.\ref{eq:linear_flux_emb},

\begin{equation}
    \label{eq:gravity_model}
    \frac{T_{ij}}{M_{i} M_{j}} \propto \exp(-d_{ij}) = - \exp(- (1 - \cos(v_i,v_j))) \propto \exp(\cos(v_i,v_j)),
\end{equation}
where $T_{ij}$ represents the \textbf{flux}, the number of people who moved between organizations $i$ and $j$,  $M_{i}$ and $M_{j}$ represent the size of an organization,  $d_{ij} = 1 - \cos(v_{i}, v_{j})$ is the embedding distance, and $\cos(v_{i}, v_{j})=\frac{v_{i} \cdot v_{j}}{\lVert v_{i} \rVert \lVert v_{j} \rVert}$  is the cosine similarity between two vectors. In the case of scientific co-affiliation, we assume all movements are without direction (symmetric), meaning that  $T_{ij} = T_{ji}$.

To rewrite  Eq.\ref{eq:gravity_model} in terms of $T_{ij}$, we replace  $M_*$ with $\mu_{i*}$, the total flux to or from $i$ which can be written as

@@

\begin{equation}
    \label{eq:skipgram_log_prob_condensed}
    \frac{1}{T}\sum_{t=1}^{T} \sum_{-c \le j \le c, j \neq 0} \log p(w_{t+j}|w_t),
\end{equation}
where the conditional probability of observing one word, $w_{O}$, given another, $w_{I}$, is defined as,

\begin{equation}
    \label{eq:skipgram_conditional_expanded}
    p(w_O|w_I) = \frac{\exp(v'_{w_O} \cdot v_{w_I})}{\sum_{w=1}^{W} \exp(v'_w \cdot v_{w_I})},
\end{equation}
where $v_w$ and $v'_w$ are ``input'' and ``output'' vector representations of each word in the vocabulary.  We use Skip-gram negative sampling model, and following the work by Levy and Goldberg \autocite{levy2014neural}, consider the model as implicitly performing matrix factorization as shown below,

\begin{equation}
    \label{eq:sgns_relation}
    A = \left(a_{ij}\right)_{W*W} = \log \left(\frac{N(i, j) |D|}{N(i) N(j)} \right) - \log k,
\end{equation}

where $N(i)\text{, } N(j) \text{ and } N(i,j)$ denotes the number of times word $i$, $j$, and word-context-pair ($i$, $j$) in the corpus respectively. Variable $k$ is the number of negative samples and $|D|$ is the size of the corpus.

From Matrix Eq.\ref{eq:sgns_relation}, we can get embedding vector with rank-d approximation by SVD, $A=L_d \Sigma_d R_d$.
Mathematically, the maximization of the objective function of the model yields $V=L_d \sqrt{\Sigma_d}$ and $V'=\sqrt{\Sigma_d}R_d$ where L and R are the left and right singular vectors of the matrix Eq.\ref{eq:sgns_relation}, respectively. Note that $V=L_d \sqrt{\Sigma_d}=\sqrt{\Sigma_d}R_d=V'$ because Eq.\ref{eq:sgns_relation} is a real symmetric matrix. With these relations, for a specific pair of words $(i ,j )$,

\begin{equation}
    \label{eq:sgns_relation_w,i, w_j}
    v_{i} \cdot v_{j} =  \log \left(\frac{N(i, j) |D|}{N(i) N(j)} \right) - \log k,
\end{equation}
where $|D|$ and $k$ are constants, such that the functional form can be re-written as,

\begin{equation}
    \label{eq:sgns_relation_w,i, w_j}
     \frac{N(i, j)}{N(i) N(j)}  \propto  \exp(v_{i} \cdot v_{j} ).
\end{equation}

In our model, $N(i)$ and $N(j)$ are total flux $M_i$ and $M_j$ respectively,  and $N(i,j)$ is just  $T_{ij}$, flux between $i$ and $j$. Then, we have the same expression with gravity model,
